---
title: "Olympiad problem. Econometrics"
output: word_document
---

Researcher John Smith estimated two models for Fertility using a data set on socio-economic indicators of Swizerland at about 1888 (47 observations): 

```{r, include=FALSE}
library("pander")
library("memisc")
library("dplyr")
mod_r <- lm(data=swiss, Fertility~Agriculture)
mod_ur <- lm(data=swiss, Fertility~Agriculture+Catholic+Education)
```

```{r, results='asis', echo=FALSE}
a <- mtable("Model 1"=mod_r, "Model 2"=mod_ur, 
       coef.style = "horizontal", 
       summary.stats=c("R-squared","N", "Deviance")) 
small_table <- data_frame(
  `Variable` = c("(Intercept)","Agriculture", "Catholic", "Education", "R^2", "RSS"), 
  `Model 1`= c("60.304 (4.251)", "0.194 (0.077)", "", "", "0.1", "6283.1"), 
  `Model 2`= c("86.225 (4.735)", "-0.203 (0.071)", "0.145 (0.030)", "-1.072 (0.156)", "0.6", "2567.9") )
colnames(small_table)[1] <- ""

small_table %>% pander()
```

Fertility - (dependent variable) standardized female fertility measure, Agriculture - percentage of males involved in agriculture, Education - percentage of educated beyond primary school for draftees, Catholic - percentage of catholic people, (standard errors are in parenthesis). 

Estimate of covariance matrix for Model 1:
```{r, echo=FALSE}
vcov(mod_r) %>% pander()
```


1. [4 pts] Find a 95\% confindence interval for coefficient before Agriculture in Model 2. 

2. [6 pts] Using appropriate test compare the two models at 5\% significance level. Clearly state the null and alternative hypothesis.

3. [10 pts] Find 95\% predictive interval for fertility in Model 1 for a region with 50\% of males involved in agriculture as occupation.


Solution:

Arithmetics errors cost [-1 pt] for each error. 

Part 1.

General formula for confidence interval [+3 pts]:

\[
[\hat{\beta} - t_{cr}se(\hat{\beta}); \hat{\beta} + t_{cr}se(\hat{\beta}) ]
\]

Critical value $t_{cr}$: either approximate (2) or from statistical table for $t_{43}$ (2.0167) or for $N(0;1)$ (1.96) --- all are considered correct [+1 pt]

Final answer (considered correct if two decimal digits after point coincide):  [-0.34; -0.06]

Part 2.

Statement of hyphothesis: [1 pt]:

$H_0$: Model 1 is true OR ($\beta_{catholic}=\beta_{education}=0$)

$H_a$: Model 1 is false, but Model 2 is true OR (at least one of $\beta_{catholic}$ or $\beta_{education}$ is not equal to zero)

Observed value of F-statistic [3 pts]

\[
F_{obs}=\frac{(R^2_2 - R^1_2)/2}{(1-R^2_2)/(47-4)} \approx 26.9
\]

Critical value for F-statistic [1 pt]

$F_{cr, 2, 43}=3.21$

Statistical conclusion [1 pt]

$H_0$ is rejected.

Part 3.

Point prediction: [+1 pt]

\[
\hat y_{n+1} = 60.304+50\cdot 0.194 = 70
\]

Estimation of standard error of regression $\hat{\sigma}^2$: [+2 pts]

\[
\hat{\sigma}^2 = \frac{RSS}{n-k}=\frac{6283}{46-2}=139.6
\]


Formula for estimate of variance of forecast: [+2 pts]
\[
se^2(\hat{\beta}_1 + \hat{\beta}_2 \cdot 50) = se^2(\hat{\beta}_1) + 50^2 se^2(\hat{\beta}_2) + 2\cdot 50\cdot \widehat{Cov}(\hat{\beta}_1, \hat{\beta}_2) \approx 2.97
\]

Formula for estimate of forecast variance [+2 pts]

\[
se^2(\hat{y}_{n+1}-y_{n+1})=se^2(\hat{\beta}_1 + \hat{\beta}_2 \cdot 50) + \hat{\sigma}^2 = 2.97 + 139.6 \approx 146
\]

Formula for predictive interval [+2 pts]

\[
[\hat{y}_{n+1} - t_{cr}se(\hat{y}_{n+1}-y_{n+1}); \hat{y}_{n+1} + t_{cr}se(\hat{y}_{n+1}-y_{n+1}) ]
\]

Critical value of t-statistic: either approximate (2) or from statistical table for $t_{45}$ (2.0141) or for $N(0;1)$ (1.96) - all are considered correct [+1 pt]

Final answer: [46; 94] (plus-minus 1 is considered ok)




